---
title: "ISA 419: Data-Driven Security"
subtitle: "09: Visualizing Data with Pandas"
author: '<br>Fadel M. Megahed, PhD <br><br> Endres Associate Professor <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("twitter"), fill = "white")` @FadelMegahed](https://twitter.com/FadelMegahed) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2024"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightSpans: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:  
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  progress = FALSE, 
  verbose = FALSE,
  dev = 'png',
  fig.height = 3,
  dpi = 300,
  fig.align = 'center')

options(htmltools.dir.version = FALSE)


miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")
if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")
if(require(urbnmapr)==FALSE) devtools::install_github('UrbanInstitute/urbnmapr')
if(require(emo)==FALSE) devtools::install_github("hadley/emo")

knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

style_mono_accent(base_color = "#84d6d3",
        base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "tachyons", "panelset", "search", "fit_screen", "editable", "clipboard"))

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,
  mute_unhighlighted_code = TRUE
)
```


## Quick Refresher of Last Class

`r emo::ji("check")` Ensure that your imported data is **technically correct** (rename columns and fix `dtypes`) 

`r emo::ji("check")` Understand how to change the unit of analysis by grouping and aggregating data.  

`r emo::ji("check")` Use the `agg()` function to do aggregations on grouped data.  



---

## Learning Objectives for Today's Class

- Create quick visualizations using the `plot` method from [pandas](https://pandas.pydata.org/docs/user_guide/visualization.html) (with an understanding of the effect of different backends).  

- Utilize `auto-viz` type plots to create a quick EDA of your data.

---
class: inverse, center, middle

# Plotting with Pandas

---

## Our Data

- We will use the `merged_ips` data set from a previous class to demonstrate how to plot data in pandas.

.font80[
```{python merged_ips}
import pandas as pd

toxic_ips = pd.read_csv(
  "https://raw.githubusercontent.com/fmegahed/isa419/main/data/listed_ip_90_all.csv", 
  header = None, names = ['ip', 'frequency', 'lastseen']
)

geolocation = pd.read_csv(
  'https://raw.githubusercontent.com/fmegahed/isa419/main/data/ip_geolocation.csv', 
  names = ['ip', 'country', 'city', 'latitude', 'longitude']
)

merged_ips = (
  toxic_ips
  .merge(right = geolocation, how = 'left', on ='ip')
  .dropna()
  .assign( lastseen = lambda df: df['lastseen'].astype('datetime64[ns]') )
)
merged_ips.dtypes[0:3]
```
]


---

## Plotting with Pandas

- The `plot` method in pandas is a wrapper around `matplotlib` (by default) and is a quick way to visualize data.

- The `plot` method is available on both `Series` and `DataFrame` objects.

![pandas_plot](https://pandas.pydata.org/docs/_images/04_plot_overview.svg)


.footnote[
<html>
<hr>
</html>

**Source:** The figure is from the [Pandas Documentation](https://pandas.pydata.org/docs/getting_started/intro_tutorials/04_plotting.html)
]


---

## Class Activity to Assess your Understanding so Far

`r countdown(minutes = 5, seconds = 0, top = 0, font_size = "2em")`

.panelset[

.panel[.panel-name[Task]
- Write Python code to produce a data frame containing the total number of toxic IP frequencies by country.  

- Then, identify the top 10 countries with the highest toxic IP frequencies.
]


.panel[.panel-name[Hints]
<html>
<details>
  <summary>Please let me know in class if you need any hints.</summary>
  <p>Use the `groupby` method to group the data by country and then use the `agg` method to aggregate the data. Then, use the `sort_values` method to sort the data in descending order by setting the `ascending` parameter to `False`.</p>
</html>
]


.panel[.panel-name[Solution]
.font80[
```{python activity_sol, echo= FALSE}
freq_countries = (
  merged_ips.groupby('country')
  .agg({'frequency': 'sum'})
  .sort_values('frequency', ascending = False)
  .head(10)
)

print(freq_countries)
```
]
]

]


---

## Plotting with Pandas (Plot `kind`)

```{r echo=FALSE}
knitr::include_url("https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.plot.html", height = "500px")
```
 

---

## Plotting with Pandas (`line` Plot)

.pull-left[
.font80[

.center[**Data Prep:**]

```{python line_plot_prep}
# Aggregating the frequencies by day
daily_freq = (
  merged_ips
  .groupby(merged_ips['lastseen'].dt.date)
  .agg(sum_freq = ('frequency', 'sum'))
  .reset_index() # to have last seen as col
  .rename(columns = {'lastseen': 'date'})
)

daily_freq.head(n=2)
```
]
]

.pull-right[
.font80[

.center[**Plotting:**]

```{python line_plot}
daily_freq.plot(
  x = 'date', y = 'sum_freq', kind = 'line',
  title = 'Toxic IP Frequencies by Day',
  xlabel = 'Date', ylabel = 'Frequency',
  figsize = (10, 4)
  )
```
]
]


---

## Plotting with Pandas (`bar` Plot)

.pull-left[

.font80[

.center[**Data:**]

```{python bar_plot_prep}
# Aggregating the frequencies by country
country_freq = (
  merged_ips
  .groupby('country')
  .agg(sum_freq = ('frequency', 'sum'))
  .sort_values('sum_freq', ascending = False)
  .head(10)
  .reset_index()
)
country_freq.head(n=2)
```
]
]


.pull-right[

.font80[
.center[**Plotting:**]

```{python bar_plot}
country_freq.plot(
  x = 'country', y = 'sum_freq', kind = 'barh',
  title = 'Top 10 Countries with Toxic IP Freq',
  xlabel = 'Country', ylabel = 'Frequency',
  figsize = (10, 6),
  color = 'red'
  )

```
]
]

---

## Plotting with Pandas (`scatter` Plot)

.pull-left[
.font80[
.center[**Data:**]

```{python scatter_plot_prep}
country_freq.head(n=5)
```
]
]

.pull-right[
.font80[
.center[**Plotting:**]

```{python scatter_plot}
country_freq.plot(
  # scatter plots are better with two numeric vars 
  # (this example is for illustration only)
  x = 'country', y = 'sum_freq', kind = 'scatter',
  title = 'Top 10 Countries with Toxic IP Freq',
  xlabel = 'Country', ylabel = 'Frequency',
  figsize = (12, 6)
  )
```
]
]


---

## Class Activity to Assess your Understanding so Far

`r countdown(minutes = 10, seconds = 0, top = 0, font_size = "2em")`

.panelset[

.panel[.panel-name[Task]

```{python sim_attack_data, include=FALSE}
import pandas as pd
import datetime as dt
import numpy as np
import random
from faker import Faker

# Set random seed for reproducibility
np.random.seed(2024)

fake = Faker()

# Predefined list of countries with top cities and their geolocations
city_geolocations = {
    'USA': [('New York', 40.7128, -74.0060), ('Los Angeles', 34.0522, -118.2437), ('Chicago', 41.8781, -87.6298)],
    'Canada': [('Toronto', 43.6510, -79.3470), ('Vancouver', 49.2827, -123.1207), ('Montreal', 45.5017, -73.5673)],
    'UK': [('London', 51.5074, -0.1278), ('Manchester', 53.4808, -2.2426), ('Edinburgh', 55.9533, -3.1883)],
    'Australia': [('Sydney', -33.8688, 151.2093), ('Melbourne', -37.8136, 144.9631), ('Brisbane', -27.4698, 153.0251)],
    'Japan': [('Tokyo', 35.6895, 139.6917), ('Osaka', 34.6937, 135.5023), ('Kyoto', 35.0116, 135.7681)]
}

# Helper functions
def generate_ip():
    return fake.ipv4()

def generate_attack_type():
    return random.choice(['DDoS', 'Phishing', 'Malware', 'Ransomware', 'SQL Injection', 'Insider Threat'])

def generate_severity():
    return random.choice(['Low', 'Medium', 'High', 'Critical'])

def generate_geo():
    country = random.choice(list(city_geolocations.keys()))
    city_info = random.choice(city_geolocations[country])
    return {"latitude": city_info[1], "longitude": city_info[2]}

# Today minus 60 days
start_date = dt.date.today() - dt.timedelta(days=45)

# Generate dataset
data = {
    "Timestamp": pd.date_range(start=start_date, periods=1000, freq='H'),
    "Source IP": [generate_ip() for _ in range(1000)],
    "Destination IP": [generate_ip() for _ in range(1000)],
    "Source Geo": [generate_geo() for _ in range(1000)],
    "Destination Geo": [generate_geo() for _ in range(1000)],
    "Attack Type": [generate_attack_type() for _ in range(1000)],
    "Severity": [generate_severity() for _ in range(1000)],
    "Attempt Count": np.random.randint(1, 20, size=1000),
    "Data Volume (MB)": np.random.randint(1, 1000, size=1000)
}

df = pd.DataFrame(data)
df["Source Latitude"] = df["Source Geo"].apply(lambda x: x["latitude"])
df["Source Longitude"] = df["Source Geo"].apply(lambda x: x["longitude"])
df["Destination Latitude"] = df["Destination Geo"].apply(lambda x: x["latitude"])
df["Destination Longitude"] = df["Destination Geo"].apply(lambda x: x["longitude"])
df.drop(columns=["Source Geo", "Destination Geo"], inplace=True)

df.to_csv("../../data/simulated_attack_data.csv", index=False)
```


- Read the [`simulated_attack_data.csv`](https://raw.githubusercontent.com/fmegahed/isa419/main/data/simulated_attack_data.csv) file into a pandas data frame.  

- Then, answer the questions in the next three tabs.  
]


.panel[.panel-name[Task 1]

- Create a histogram of the `Attempt Count` variable.

- What does the histogram tell you about the `Attempt Count` variable?

```{python sim_hist, include=FALSE}
sim_attack_df = pd.read_csv("../../data/simulated_attack_data.csv")

sim_attack_df.plot(
  y = 'Attempt Count', kind = 'hist',
  title = 'Histogram of Attempt Count',
  xlabel = 'Attempt Count', ylabel = 'Frequency',
  figsize = (10, 6),
  color = 'red'
  )
```

.can-edit.key-activity12a[
- Edit me to answer the question above.
]
]

.panel[.panel-name[Task 2]
 - Create a scatter plot of the `Source Latitude` and `Source Longitude` variables.  
 
]


.panel[.panel-name[Task 3]
 - Utilize [this Stack Overflow thread](https://stackoverflow.com/questions/53233228/plot-latitude-longitude-from-csv-in-python-3-6) to convert the scatter plot of the `Source Latitude` and `Source Longitude` variables into an interactive symbols map. 

```{python, echo=FALSE}
import plotly.express as px

fig = px.scatter_geo(sim_attack_df, lat="Source Latitude", lon="Source Longitude")
fig.show()
```
]
]


---
class: inverse, center, middle

# Automated Viasualizations in Python

---

## The `ydata-profiling` Package

Data quality profiling and exploratory data analysis (EDA) are crucial steps in any business analytics application.   

[ydata-profiling](https://docs.profiling.ydata.ai/latest/) automates and standardizes the generation of detailed reports, complete with statistics and visualizations. 

The significance of the package lies in how it streamlines the process of understanding and preparing data for analysis in a single line of code!

---

## Usage of `ydata-profiling`

```{python profiling_example, results='hide'}
import pandas as pd
from ydata_profiling import ProfileReport #<<

profile = ProfileReport(sim_attack_df, title="Pandas Profiling Report", explorative=True) #<<

# the next line is needed since I am not using Colab for making the slides
profile.to_file("../../figures/sim_attack_data_report.html") 
```


---

## Output of `ydata-profiling`

```{r echo=FALSE}
knitr::include_url("https://fmegahed.github.io/isa419/spring2024/class09/sim_attack_data_report.html", height = "500px")
```


---

## Class Activity

`r countdown(minutes = 10, seconds = 0, top = 0, font_size = "2em")`

.panelset[

.panel[.panel-name[Task]

- Identify a similar package to `ydata-profiling` in Python.  

- Then, use the package to generate a report for the `merged_ips` data set.  

- Share the report with your neighboring classmate.  

- Discuss the insights and visualizations in your approach(es).  

]
]

---

class: inverse, center, middle

# Recap

---

## Summary of Main Points

By now, you should be able to do the following:  

- Create quick visualizations using the `plot` method from [pandas](https://pandas.pydata.org/docs/user_guide/visualization.html) (with an understanding of the effect of different backends).  

- Utilize `auto-viz` type plots to create a quick EDA of your data.

---

## ðŸ“ Review and Clarification ðŸ“

1. **Class Notes**: Take some time to revisit your class notes for key insights and concepts.
2. **Zoom Recording**: The recording of today's class will be made available on Canvas approximately 3-4 hours after the end of class.
3. **Questions**: Please don't hesitate to ask for clarification on any topics discussed in class. It's crucial not to let questions accumulate. 

