---
title: "ISA 480: Data Driven Security"
subtitle: "04 - An Overiew of ETL Operations"
author: Fadel M. Megahed
institute: |
  | Endres Associate Professor
  | Department of Information Systems and Analytics
  | Farmer School of Business
  | Miami University
  | Email: fmegahed@miamioh.edu
  | Office Hours: [Automated Scheduler for Office Hours](https://calendly.com/fmegahed)
  | 
date: "Fall 2022"
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    includes: 
      in_header: ../../style_files/structure.txt
latex_engine: xelatex
urlcolor: links
linkcolor: links
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      cache = TRUE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5,
                      fig.height= 2.5,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra)
```

# Preface

### Quick Recap of Classes 01-03
\begin{block}{\textbf{Learning Objectives Discussed}}
  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Understand the structure and expectations of this course} \\
  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Define information security and its main goals} \\
  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Understand that breaches are frequent and target different industries} \\
  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Understand the basic operations in either R or Python}
\end{block}


### Plan for Today's Class
\begin{block}{\textbf{Learning Objectives}} 
  \begin{itemize}
  \item \textbf{Describe the data analytics process}
  \item  \textbf{Understand the importance of ETL and data preprocessing}
  \begin{itemize}
      \item \textbf{Explain what do we mean by \textcolor{miamired}{technically correct data}}
      \item \textbf{Explain what do we mean by \textcolor{miamired}{consistent data}}
  \end{itemize}
  \item  \textbf{Perform basic ETL in R and/or Python}
  \end{itemize}
\end{block}


# The Data Analytic Process


### Big Picture: Levels of Data Analytics

\centering{\href{http://www.ithappens.nu/levels-of-data-analytics/}{\includegraphics[width=\textwidth, height=0.8\textheight, keepaspectratio, frame]{../../figures/levelsDA.png}}}


### Implementation Frameworks for ML/Data Analytics

\begin{table}[h]
  \centering
  \begin{tabular}{|l|l|l|}
  \hline
      \rowcolor{miamired}
       \textbf{\textcolor{white}{The KDD Process}} & \textbf{\textcolor{white}{SEMMA}} & \textbf{\textcolor{white}{CRISP-DM}} \\ \hline \hline 
       \rowcolor{gray!30}
       Pre KDD & ------------- & Business understanding \\ \hline \hline 
       Selection & Sample & \multirow{2}{*}{Data understanding} \\ \cline{1-2}
       Pre processing & Explore & \\ \hline \hline 
       \rowcolor{gray!30}
       Transformation & Modify & Data preparation \\ \hline \hline 
       Data mining & Model & Modeling \\ \hline \hline 
       \rowcolor{gray!30}
       Interpretation/Evaluation & Assessment & Evaluation \\ \hline \hline  
       Post KDD &  ------------- & Deployment \\ \hline \hline 
  \end{tabular}
\end{table}

\centering{Table: A comparison of the three most commonly used Data Analytic/ Machine Learning Frameworks. Table adapted from} @azevedo2008kdd.


### Comment on the three Implementation Frameworks
\begin{block}{\textbf{Practical Issue with Existing Guidance}}
    \textbf{These three frameworks encourage iterating, but do not provide sufficient guidance on how.}
\end{block}

\centering 

\href{https://scikit-learn.org/stable/tutorial/machine_learning_map/index.html}{\includegraphics[width=\textwidth, height=0.54\textheight, keepaspectratio, frame]{../../figures/ml_map.png}}
    
\footnotesize \textbf{Source:} SciKit-Learn @scikit2020choosing.


#  Important ETL Concepts

### Data Analysis: A Crowd Sourced Definition from Wikipedia

Wikipedia @wikipedia2020data defines \textcolor{miamired}{\textbf{data analysis}} to be the process of: 

\vspace{0.25\baselineskip}

\begin{quotation}
\noindent inspecting, cleansing, transforming and modeling data with the goal of discovering useful information, informing conclusion and supporting decision-making. Data analysis has multiple facets and approaches, encompassing diverse techniques under a variety of names, and is used in different business, science, and social science domains.
\end{quotation}


### Data Analysis Value Chain

\centering

\href{https://cran.r-project.org/doc/contrib/de_Jonge+van_der_Loo-Introduction_to_data_cleaning_with_R.pdf}{\includegraphics[width=\textwidth, height=0.77\textheight, keepaspectratio, frame]{../../figures/statValueChain.PNG}}

\vspace{-0.5\baselineskip}

\footnotesize \textbf{Source:} The figure is from @de2013introduction. Please click on the image and read p. 7-8 from the reference.

### Useful Operations/Functions (in R/Python)

\begin{table}[h]
\centering
  \begin{tabular}{|p{0.2\textwidth}|p{0.35\textwidth}|p{0.35\textwidth}|}
  \hline
      \rowcolor{miamired}
       \textbf{\textcolor{white}{Goal}} & \textbf{\textcolor{white}{R}} & \textbf{\textcolor{white}{pandas}} \\ \hline \hline 
       \rowcolor{gray!30}
       structure/classes of variables & \texttt{str()} \newline \texttt{dplyr::glimpse()} & \texttt{df.info()} \\
       summaries for each col/variable & \texttt{summary()} & \texttt{df.describe()} \\
       \rowcolor{gray!30}
       dimensions of object & \texttt{dim()} & \texttt{df.shape} \\
       convert column types & \texttt{df\$x1 = as.numeric(df\$x1)} \newline \texttt{df \%>\%} \newline \textcolor{white}{hh} \texttt{mutate(x1 = as.numeric(x1))} & \texttt{df["x1"] = df["x1"].astype(float)}  \newline \texttt{df = df.astype(\{"x2": int, "x3": complex\})} \\ 
       \rowcolor{gray!30} 
       subset data frame by row numbers & \texttt{slice(df, 1:10)} & \texttt{df.iloc[:9]} \\ \hline \hline
    \end{tabular}
\end{table}

See the \href{https://pandas.pydata.org/pandas-docs/stable/getting_started/comparison/comparison_with_r.html}{Panda's Official Comparison with R Libraries} for more details.


### A Primer on Technically Correct and Consistent Data

Let us discuss the concepts of **technically correct** and **consistent** data based on the `KDDTrain+.csv` dataset. Note that the data was created by @tavallaee2009detailed (click [here](https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=5356528) to access their paper).

**Comments:**   

- I have intentionally picked a large dataset so we can think about how we can design a process to check for technically correct and consistent data.  
- How many rows, variables and unique values for the `label` column do we have?   
- Can you report summary statistics for the data?   
- How can we visualize such data quickly in both R (`DataExplorer::create_report()`) and Python (`pandas-profiling`)?  


```{r kddNSL, include = FALSE}
pacman::p_load(tidyverse, magrittr)
df = read_csv("../../data/KDDTrain+.csv")
```



# ETL Operations on IP Addresses

### Developing an intuition {.allowframebreaks}

```{r cmd}
system("ping google.com", intern = T)
```


\begin{block}{\textbf{Dissecting the IP Address}}
  \begin{itemize}
    \item IP addresses is one of the most fundamental security domain elements
    \item IPV4 is comprised of \textcolor{miamired}{four bytes}, and is commonly stored in a \textcolor{miamired}{dotted-decimal notation}
      \begin{itemize}
        \item A byte (i.e. 8-bits) can range in values from 0 to 255
        \item The total number of IPv4 addresses is equal to: $\mathbf{(2^8)^4 = 2^{32} = 4,294,967,296}$
        \item Thus, any ip address can be converted to/from a 32-bit integer value
      \end{itemize}
    \item The conversion from a 32-bit value can be used to save space since each string character is stored in one byte
    \begin{itemize}
      \item Thus, the storage cost is reduced from $\mathbf{15\times 8 = 120}$ to $\mathbf{4 \times 8 =32}$ bits per IP address (assuming most efficient storage for both object types)
    \end{itemize}
  \end{itemize}
\end{block}


\centering{\href{https://www.wikihow.com/Convert-from-Binary-to-Decimal}{\includegraphics[width=\linewidth, height=0.8\textheight,frame, keepaspectratio]{../../figures/binaryToDecimal.png}}}

### Converting the dotted-decimal notation IP to a 32-Bit Integer {.allowframebreaks} 
To understand how the binary notation can be extended to IP addresses, let us convert the following IP addresses to their 32bit Integer number.

\begin{itemize}
  \item Q1: "0.0.0.1" %Answer: 1

  \item Q2: "0.0.0.3" %Answer: 3

  \item Q3: "0.0.1.0" %Answer: 256

  \item Q4: "0.0.1.240" %Answer: 496

  \item Q5: "0.0.100.240" %Answer: 25840

\end{itemize}

You can check your answer using the [iptools package](https://cran.r-project.org/web/packages/iptools/iptools.pdf) function titled: `ip_to_numeric`.

```{r iptonumeric}
print("Conversion of IP addresses using R")
pacman::p_load(iptools)

ip_to_numeric("0.0.100.240")

ip_to_numeric("0.0.100.240") %>% numeric_to_ip()

```


```{python ipConversion}
print('Python Conversion of IP addresses')

import ipaddress

int(ipaddress.ip_address('0.0.100.240'))

str(ipaddress.ip_address(25840))

```


### Classless Inter-Domain Routing (CIDR) Blocks/ Ranges {.allowframebreaks}

\centering{\href{https://en.wikipedia.org/wiki/IPv4_address_exhaustion}{\includegraphics[width=\linewidth, height=0.8\textheight,frame, keepaspectratio]{../../figures/IPv4.png}}}


\centering{\href{https://myip.ms/info/whois/104.20.31.157/k/649487422/website/filgoal.com}{\includegraphics[width=\linewidth, height=0.8\textheight,frame, keepaspectratio]{../../figures/cidr-filgoal.PNG}}}

Number of IP addresses in picture:  $= 2^{32-12} = 2^{20} = (31-16+1) * 2^8 * 2^8 = 1,048,576$

```{r cidrs, eval=FALSE}
iptools::ips_in_cidrs("104.16.244.12","104.16.0.0/12") # R
```

```{python cidrs2}
import ipaddress as ip
ip.ip_address('104.16.244.12') in ip.ip_network('104.16.0.0/12') # Python
```


### Live Demo: Dissecting IP addresses

We will explore working with IP addresses. Specifically, we will work with IP addresses since they are commonly used in Info Security to highlight suspicious domains. Please check @att2020reputation and @sans2020suspicious for more details. 

By the end of the demo, you should be able to do the following:
\begin{itemize}
  \item Explain the basics of the IPV4 structure
  \item Convert IP addresses to their corresponding 32-bit integer representation
  \item Be able to use online APIs to extract relevant information for a given API/ list of APIs
\end{itemize}

\textcolor{miamired}{\textbf{Prior to our live demo, let us examine some online Miami University IP address information to introduce you to the IP Stack API.}} You will be required to create an account and utilize the \textcolor{miamired}{\underline{\textbf{free}}} API.

```{r ips1, include=FALSE, eval=FALSE}
source("ipstackKey.R") # where I saved my key
pacman::p_load(rvest, tidyverse, magrittr, jsonlite) # loading needed packages
miamiURLs = 
  read_html("https://tools.tracemyip.org/search--isp/miami+university") %>% 
  html_node("#tlzRDTIPv4") %>% html_table() %>% select(c(2)) %>% unlist()

url_base = "http://api.ipstack.com/"

dfLocations = as.data.frame(matrix(nrow=50, ncol = 8))
apiFields = "latitude,longitude,zip,city,region_name,type,ip"
colnames(dfLocations) = strsplit(apiFields, ",") %>% unlist()

for (i in 1:6) {
  dataFromAPI = fromJSON(
    paste0(url_base, miamiURLs[i], "?access_key=", accessKey,
           "&fields=",apiFields))
  temp = unlist(dataFromAPI)
  dfLocations[i,] = temp
}

dfLocations
```


# Recap

### What I have tried to accomplish in today's class
\begin{block}{\textbf{Learning Objectives}} 
  \begin{itemize}
  \item \textbf{Describe the data analytics process}
  \item  \textbf{Understand the importance of ETL and data preprocessing}
  \begin{itemize}
      \item \textbf{Explain what do we mean by \textcolor{miamired}{technically correct data}}
      \item \textbf{Explain what do we mean by \textcolor{miamired}{consistent data}}
  \end{itemize}
  \item  \textbf{Perform basic ETL in R and/or Python}
  \end{itemize}
\end{block}



### References {.allowframebreaks}
::: {#refs}
:::

---

\maketitle
