---
title: "ISA 419: Data Driven Security"
subtitle: "08 - Descriptive Analytics (Cont.)"
author: Fadel M. Megahed
institute: |
  | Endres Associate Professor
  | Department of Information Systems and Analytics
  | Farmer School of Business
  | Miami University
  | Email: fmegahed@miamioh.edu
  | Office Hours: [Automated Scheduler for Office Hours](https://calendly.com/fmegahed)
  | 
date: "Fall 2022"
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    includes: 
      in_header: ../../style_files/structure.txt
latex_engine: xelatex
urlcolor: links
linkcolor: links
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5,
                      fig.height= 2.5,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, tidyverse, xtable)
```

# Recap 

### Our Course Progress so Far

\begin{block}{\textbf{Learning Objectives Discussed}}
  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Define information security and its main goals} \\
  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Understand that breaches are frequent and target different industries} \\
  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Understand the basic operations in either R or Python} \\
   $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Describe the data analytics process} \\
  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Describe/Convert  ETL data to be both technically correct and consistent} \\
  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Summarize and examine data through ETL operations}
\end{block}


### Security Datasets that you have Examined So Far

- [Have I been pawned?](https://haveibeenpwned.com/)    

- [Los Alamos National Lab's User Authentication Dataset](https://csr.lanl.gov/data/auth/)  

- [NSL-KDD Cup 1999 Intrusion Dataset](https://www.unb.ca/cic/datasets/nsl.html)  

- [Allien Vault IP Reputation Dataset](https://reputation.alienvault.com/reputation.data)

### Learning Objectives for Today's Class

- Describe impact of data types on how we summarize them  

- Intro to Data Encoding and Visualizations  



# Data Models

### On the Theory of Scales of Measurement

\centering {\includegraphics[width=\textwidth, height=1.5\textheight, keepaspectratio,frame]{../../figures/stevens_science_paper.png}}


### Data Types (from S. Stevens, Theory of Scales)

\centering {\includegraphics[width=\textwidth, height=0.85\textheight, keepaspectratio,frame]{../../figures/datatypes.png}}


### Nnominal, ordinal, or numeric for Portmap.csv?

```{r portmap_types, echo=FALSE}
pacman::p_load(tidyverse, magrittr)
df = read.csv("../../data/Portmap.csv", nrow=1000, stringsAsFactors = T)
df %<>% select('Total.Fwd.Packets', 'Total.Backward.Packets', 
               'Max.Packet.Length', 'Average.Packet.Size', 'Label')
colnames(df) = c("FWD", "Back", "PL", "PS", "Label")
glimpse(df)
```



# Descriptive Analytics/ Exploratory Data Analysis (EDA)

### Objectives

\textbf{Our primary goal is to \textcolor{miamired}{develop a better understanding of the data.} The following guidelines from} **@rds2016** \textbf{can be useful:}
\begin{itemize}
  \item Use questions as tools to guide your investigation.
  \begin{itemize}
    \item  When you ask a question, it focuses your attention on a specific part of your dataset and helps you decide which graphs, models, or transformations to make.
  \end{itemize}
  \item EDA is fundamentally a creative process, where you should ask \textcolor{miamired}{quality} questions through asking a large quantity of questions.
  \item There is no rule about which questions you should ask to guide your research. However, two types of questions will always be useful for making discoveries within your data.
  \begin{itemize}
    \item What type of variation occurs within my variables?
    \item What type of covariation occurs between my variables?
  \end{itemize}
\end{itemize}

### Implementation Strategies

**In my estimation, the EDA process typically combines:**

- **Descriptive statistics/modeling,** where statistical techniques for summarizing and describing the variation within the data are used.  
- **Visualization techniques,** charting the data allows us to get a different perspective of the data.  

### A Synthetic Example: The Anscombe Dataset {.allowframebreaks}

**In a seminal paper, @anscombe1973graphs stated:** *Few of us escape being indoctrinated with these notions* 

- *numerical calculations are exact, but graphs are rough;*  
- *for any particular kind of statistical data there is just one set of calculations constituting a correct statistical analysis;*  
- *performing intricate calculations is virtuous, whereas actually looking at the data is cheating.*

**He proceeded by stating that** *a computer should make both calculations and graphs. Both sorts of output should be studied; each will contribute to understanding.*  

**Now, let us consider his four datasets, each consisting of eleven (x,y) pairs.**



```{r anscombe1}
anscombe %>% xtable::xtable() -> xt
align(xt) <- "lrrrrrrrr"

cat("---")
```

```{r anscombe2, results='asis'}
print(xt, include.rownames=F) 
```
 
```{r anscombe3, results='asis'}
pacman::p_load(Tmisc) # same data but in 3 columns
df = quartet 
df %>% group_by(set) %>% 
  summarise(x.mean = mean(x), x.sd = sd(x),
            y.mean = mean(y), y.sd = sd(y),
            corr = cor(x, y)) %>% kable(digits = 3)
```
The code in this slide and next is based on <https://rpubs.com/turnersd/anscombe>. 

```{r anscombe4, results='asis', fig.height=3}
ggplot(quartet, aes(x, y)) + geom_point() + 
  geom_smooth(method = lm, se = FALSE) + facet_wrap(~set)
```

### Anscombe-Like Mistakes in Research and Practice
In my estimation, Figure 8c from @cai2012neuronal represents an example where regression should not have been performed. 

\centering \href{https://www.jneurosci.org/content/32/11/3791/tab-figures-data}{\includegraphics[width=\textwidth, height=0.7\textheight, keepaspectratio,frame]{../../figures/F8Large.jpg}}

### Some Useful Statistical Summaries {.allowframebreaks}

In lab 01 we have described several variables, you may be interested in performing some statistical summaries for a grouping variable. 

```{r portmap, results='asis'}
pacman::p_load(magrittr)
df = read.csv("../../data/Portmap.csv", nrow=1000, stringsAsFactors = T)
df %<>% select('Total.Fwd.Packets', 'Total.Backward.Packets', 
               'Max.Packet.Length', 'Average.Packet.Size', 'Label')
colnames(df) = c("FWD", "Back", "PL", "PS", "Label")

df %>% group_by(Label) %>% 
  summarise_all(list(med= median, avg = mean, sd= sd)) %>% 
  kable() %>% 
  kable_styling(position = "center", latex_options = "scale_down")

```

### Some Useful Statistical Tests

For a more detailed statistical analysis of your data. I recommend the table in <https://stats.idre.ucla.edu/other/mult-pkg/whatstat/>, which provides general guidelines for choosing a statistical test. \textcolor{miamired}{To remind you to examine the link, I included the first 6 rows (limited to the first 4 columns) of that table below.}  

```{r tablefromUCLA, echo=F, cache=TRUE, results='asis'}
pacman::p_load(tidyverse, rvest)

"https://stats.idre.ucla.edu/other/mult-pkg/whatstat/" %>% read_html() %>% 
  html_node("#genesis-content > article > div > table") %>% html_table() %>% 
  select(c(1:4)) %>%  xtable() -> ucla
colnames(ucla) = c('no. Ys', 'type of Xs', 'type of Ys', 'Test(s)')

align(ucla) = "lp{0.42in}p{1.38in}p{1.38in}p{1.38in}"

print(head(ucla), include.rownames=F) 
```

### Fundamentals of Data Viz: Guidelines 

\begin{block}{\textbf{The Visual Display of Quantitative Data (E. Tufte)}}
		\begin{itemize}
			\setbeamercovered{transparent}
			\item \textcolor{miamired}{\textbf{Show the data}}
			\item \textbf{Lead to thinking about the \textcolor{miamired}{substance} rather than  something else}
			\item \textbf{\textcolor{miamired}{Avoid distorting} what the data have to say}
			\item \textbf{\textcolor{miamired}{Present many numbers in a small space}}
			\item \textbf{\textcolor{miamired}{Make large datasets coherent}}
			\item \textbf{\textcolor{miamired}{Encourage the eye to compare different pieces of the data}}
			\item \textbf{\textcolor{miamired}{Reveal the data at several levels of detail}, from a broad overview to the fine structure}
			\item \textbf{\textcolor{miamired}{Serve a purpose:} description, exploration, tabulation, decoration}
			\item \textbf{\textcolor{miamired}{Be closely integrated with the statistical and verbal descriptions of a data set}}
		\end{itemize}
	\end{block}
	\begin{center}
		\tiny Source: From Tufte, E. R. (2001). The visual display of quantitative information. Cheshire, Conn: Graphics Press, P. 13.
		
	\end{center}

### Fundamentals of Data Viz: Data Types + Encoding

\begin{center}
  \href{https://dl.acm.org/doi/pdf/10.1145/22949.22950?casa_token=ribMBBRoenoAAAAA:rg1RxdU0rR91hbGsYjaL0m_TIYMI8WrrrJ58O0W6qJFo8kwawO1nKJLkkBfcD0EQioX6xuFHL5Ug}{\includegraphics[height=0.75\textheight, frame]{../../figures/mackinlay.png}} 
\end{center}
Source: Please refer to @mackinlay1986automating for more details on proper and automated data encoding.

### Fundamentals of Data Viz: Grammar of Graphics {.allowframebreaks}


In order to create a plot, you:

  1. Call the ggplot() function which creates a blank canvas  
  2. Specify **aesthetic mappings**, which specifies how you want to map variables to visual aspects. In this case we are simply mapping the *Label* and *Average.Packet.Size* variables to the x- and y-axes.  
  3. You then add new layers that are geometric objects which will show up on the plot. In this case we add `geom_point()` to add a layer with points (dot) elements as the geometric shapes to represent the data.
  
```{r portMapplot, fig.width=6}
df = read.csv("../../data/Portmap.csv", nrow=1000, stringsAsFactors = T)
df %<>% select ('Flow.ID','Source.IP','Destination.IP',
                'Timestamp', 'Flow.Duration', 'Total.Fwd.Packets',
                'Total.Backward.Packets', 'Max.Packet.Length',
                'Average.Packet.Size', 'Active.Mean', 
                'SimillarHTTP', 'Label')

# create canvas
ggplot(df) +
  ggtitle("Creating an Empty Canvas")

# variables of interest mapped
ggplot(df, aes(x=Label, y=Average.Packet.Size)) + 
  ggtitle("Canvas with variables and no data")

# data plotted: dot plot
ggplot(df, aes(x=Label, y=Average.Packet.Size)) + geom_point() + 
  ggtitle("Scatter/Dot Plot: Canvas with variables and data")

# data plotted: bar
ggplot(df, aes(x=Label, y=Average.Packet.Size)) + 
  geom_bar(stat="identity") + 
  ggtitle("Bar Plot: Canvas with variables and data")

# data plotted: hist
ggplot(df, aes(x=Average.Packet.Size)) + geom_histogram(bins=20) +
  ggtitle("Histogram: Canvas with variables and data")

# Some nice functions from the Data Explorer Package
pacman::p_load(DataExplorer)

plot_boxplot(df, by="Label", ncol=3) # fav plot

plot_bar(df[1:20,c('Source.IP','Destination.IP')],
         title = "Plot's Quality is Impacted by Values of Strings",
         ncol = 2) # first twenty rows and two factor/string cols

plot_histogram(df, ncol=3)
```



# Recap

### Things to Do {.allowframebreaks}

- You are highly encouraged to go through the following materials: (see refs for link) 
  - **EDA:** Chapter 7 in @rds2016  
  - **Data Viz:** Chapters 2 and 5 in @wilke2019fundamentals
  - **ggplot2 Tutorial:** Chapters 2 and 10 in @wickham2020gg. **You should also bookmark the following references**
    - *ggplot2-book:* @wickham2020gg is freely available at <https://ggplot2-book.org/>
    - *R cheatsheets:* <https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf>
  

### References {.allowframebreaks}
::: {#refs}
:::

---

\maketitle