---
title: "ISA 419: Data-Driven Security"
subtitle: "02: Python Programming Basics"
author: '<br>Fadel M. Megahed, PhD <br><br> Endres Associate Professor <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("twitter"), fill = "white")` @FadelMegahed](https://twitter.com/FadelMegahed) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2024"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightSpans: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:  
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE, 
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 3,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)


miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")
if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")
if(require(urbnmapr)==FALSE) devtools::install_github('UrbanInstitute/urbnmapr')
if(require(emo)==FALSE) devtools::install_github("hadley/emo")

knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "tachyons", "panelset", "search", "fit_screen", "editable", "clipboard"))

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,
  mute_unhighlighted_code = TRUE
)
```


## Quick Refresher of Last Class

`r emo::ji("check")` Define **information security (Infosec)**, its **main goals**, and how it fits within a firm’s overall security protocols  

`r emo::ji("check")` Describe the **three main steps** in information security: prevention, detection, and response.

`r emo::ji("check")` Explain why **prevention as a sole security measure is deemed to fail**. 

`r emo::ji("check")` Describe **course structure**, **goals**, and **overview**.  


---

## Learning Objectives for Today's Class

- Use pseudocode to map out a problem.  

- Python Syntax, Data Types, and Data Structures  

- Convert data types using type casting  

- Manipulate lists and use methods on lists


---
class: inverse, center, middle

# Use Pseudocode to Map out a Cybersecurity Problem

---

## Background: (Distributed) Denial of Service Attacks

<center>
<iframe width="810" height="472.5" src="https://www.youtube-nocookie.com/embed/bDAY-oUP0DQ?si=9RiVqrFSJdStTIeS" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</center>


.footnote[
<html>
<hr>
</html>

**Credits:** Jeff Crume. (2023) [IBM Technology: Denial of Service Attacks Explained](https://www.youtube.com/watch?v=bDAY-oUP0DQ).
]

---

## Simplifying our Task Ahead: SYN Flood Attacks

- We will focus on a **SYN flood attack**, a form of DDoS attack.
- Attackers send a succession of SYN requests to a target's system.



---


```{python pseudo_code, eval=FALSE}
BEGIN DDoS_Detection
    `SET` threshold_limit `to` a significant number representing unusual traffic
    `SET` time_window `to` the period of monitoring traffic (e.g., per minute)
    `SET` ip_request_count `to` an empty data structure (e.g., dictionary)

    for each request in network_traffic
        `GET` ip_address from request
        `GET` request_time from request

        if ip_address is not in ip_request_count
            `SET` ip_request_count[ip_address] to 1
        else
            `INCREMENT` ip_request_count[ip_address] by 1

        if ip_request_count[ip_address] exceeds threshold_limit within time_window
            `FLAG` ip_address as potential DDoS source
            `TAKE` necessary action (e.g., block IP, alert admin)

        if time elapsed > time_window
            `RESET` ip_request_count for each ip_address

    `END` FOR loop
END DDoS_Detection
```



```{python, echo = FALSE, eval=FALSE}
import random
import time

# Configuration
threshold_limit = 100  # Threshold of connections per time window
time_window = 60  # Time window in seconds
monitoring_period = 300  # Total monitoring period in seconds
regular_ips = [f"192.168.1.{i}" for i in range(2, 255)] # Regular traffic IPs
attacker_ip = "192.168.1.1" # Attacker's IP generating excessive traffic

# Simulated data
network_traffic = []
for _ in range(500):  # Total regular requests
    network_traffic.append({
        'ip_address': random.choice(regular_ips),
        'timestamp': time.time() + random.randint(1, 300)
    })

# Inject attack traffic from one IP
for _ in range(200):  # Total attack requests
    network_traffic.append({
        'ip_address': attacker_ip,
        'timestamp': time.time() + random.randint(1, 300)
    })

# Shuffle the traffic to simulate a more realistic scenario
random.shuffle(network_traffic)

# Monitoring function
def detect_ddos(traffic, threshold, window, period):
    start_time = time.time()
    ip_request_count = {}

    for request in traffic:
        current_time = time.time()
        if current_time - start_time > period:
            break

        ip_address = request['ip_address']
        timestamp = request['timestamp']

        # Increment the count for the IP address
        ip_request_count[ip_address] = ip_request_count.get(ip_address, 0) + 1

        # Check if the threshold is exceeded within the time window
        if ip_request_count[ip_address] == threshold:
            time_first_request = min(timestamp for req in traffic if req['ip_address'] == ip_address)
            if current_time - time_first_request <= window:
                print(f"Potential DDoS attack detected from {ip_address}")

        # Reset counters for IP addresses not seen within the time window
        ip_request_count = {ip: count for ip, count in ip_request_count.items() if current_time - timestamp <= window}

# Run the DDoS detection
detect_ddos(network_traffic, threshold_limit, time_window, monitoring_period)

```

