---
title: "ISA 419: Data Driven Security"
subtitle: "16 - Intro to Insider Threat and User Behavior Analysis"
author: Fadel M. Megahed
institute: |
  | Endres Associate Professor
  | Department of Information Systems and Analytics
  | Farmer School of Business
  | Miami University
  | Email: fmegahed@miamioh.edu
  | Office Hours: [Automated Scheduler for Office Hours](https://calendly.com/fmegahed)
  | 
date: "Fall 2022"
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    includes: 
      in_header: ../../style_files/structure.txt
latex_engine: xelatex
urlcolor: links
linkcolor: links
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5,
                      fig.height= 2.5,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, tidyverse, xtable)
```

### Learning Objectives for Today's Class

\begin{block}{\textbf{Objectives}}
  \begin{itemize}
    \item Understand the relationship between useability and user authentication
    \item Describe different types of insider attacks
    \item Describe general approaches to defending insider threats/attacks
    \item Examine a real-world dataset on user profiling
  \end{itemize}
\end{block}

# A Primer on User Authentication

### Preface

**So far in class, we have examined the following datasets that relate to the concept of user authentication.** \textbf{\color{miamired}{Q: What were these datasets about?}}

- \textbf{\href{https://haveibeenpwned.com/}{Have you been pawned?}}

- \textbf{\href{https://figshare.com/articles/Yahoo_Password_Frequency_Corpus/2057937}{Yahoo Password Frequency Corpus}}

- \textbf{\href{https://csr.lanl.gov/data/auth/}{Los Alamos National Lab - User-Computer Authentication Associations in Time}}


### Authentication: How does that fit in the CIA-triad?
\begin{block}{\textbf{Confidentiality, Integrity, and Availability (CIA)}}
		\begin{itemize}
			\item \textbf{\textcolor{miamired}{Confidentiality:} Preventing unauthorized reading/disclosure of information.}
			\begin{itemize}
			  \item \textbf{Tools:} \textcolor{darkgreen}{User authentication}, data encryption, ...
			\end{itemize}
			
			\vspace{\baselineskip}
			
			\item \textbf{\textcolor{miamired}{Integrity:} Preventing unauthorized modification of information.}
			\begin{itemize}
			  \item \textbf{Tools:} Cryptographic hashing, digital signature, ...
			\end{itemize}
			
			\vspace{\baselineskip}
			
			\item \textbf{\textcolor{miamired}{Availability:} Preventing unauthorized withholding of information/ resources.}
			\begin{itemize}
			  \item \textbf{Tools:} Intrusion detection, distributed service, ...
			\end{itemize}
		\end{itemize}
	\end{block}


### Insider Threats: Unauthorized Disclosure of Information [1]

\begin{figure}
  \centering
  \includegraphics[height=0.65\textheight]{../../figures/Edward_Snowden-2.jpg}
\end{figure}

**Whether you call him ``a traitor or a whistle-blower, he earned one label about which there's no debate: insider threat.''** @lawrence2015


### Insider Threats: Unauthorized Disclosure of Information [2]

- **[Tesla:](https://www.forbes.com/sites/forbestechcouncil/2018/07/19/what-teslas-spygate-teaches-us-about-insider-threats/#1e4574de5afe)** A malicious insider sabotaged systems and sent proprietary data to third parties

- **[Facebook:](https://www.nbcnews.com/tech/social-media/facebook-investigating-claim-engineer-used-access-stalk-women-n870526)** A security engineer abused his access to stalk women.  

- **[CocaCola:](https://www.bleepingcomputer.com/news/security/coca-cola-suffers-breach-at-the-hands-of-former-employee/)**  A malicious insider stole a hard drive full of personnel data.  

- **[Suntrust Bank:](https://www.darkreading.com/attacks-breaches/suntrust-ex-employee-may-have-stolen-data-on-15-million-bank-clients/d/d-id/1331610)** A malicious insider stole personal data, including account information, for 1.5 million customers to provide to a criminal organization.

\vspace{5\baselineskip}

Source: Slide and references are based on @peters2019. Please read the news/blog articles.


### Frequency of Insider Threats [1]

``Despite some variation from year to year, inside jobs occur about as often as outside jobs. The lesson here, though, surely is as simple as this: organizations have to anticipate attacks from all quarters.'' CSI/FBI COMPUTER CRIME AND SECURITY SURVEY 2005

Source: Slide is from CS259D, Stanford University, \url{https://web.stanford.edu/class/cs259d/lectures/Session3.pdf}

### Frequency of Insider Threats: Verizon DBIR 2019 [2]

\begin{figure}
  \centering
  \href{https://enterprise.verizon.com/resources/reports/dbir/2019/results-and-analysis/}{\includegraphics[height=0.85\textheight, width=\textwidth, keepaspectratio, trim={0in 3in 0in 2in}, clip, frame]{../../figures/results-and-analysis-figure6.pdf}}
\end{figure}



### Frequency of Insider Threats: Verizon DBIR 2019 [3]

\begin{figure}
  \centering
  \href{https://enterprise.verizon.com/resources/reports/dbir/2019/summary-of-findings/}{\includegraphics[height=0.85\textheight, width=\textwidth, keepaspectratio, trim={0.1in 0.9in 1in 0.4in}, clip, frame]{../../figures/introduction-figure4.pdf}}
\end{figure}


### Suggested Reading: The New Yorker Article

\begin{figure}
  \centering
  \href{https://www.newyorker.com/news/news-desk/how-the-f-b-i-cracked-a-chinese-spy-ring}{\includegraphics[height=0.85\textheight, width=\textwidth, keepaspectratio, frame]{../../figures/newyorker.JPG}}
\end{figure}



# Types of Insider Attacks

### Types of Insider Attacks

- **Unauthorized extraction, duplication, or exfiltration of data**  

- **Tampering with data (unauthorized changes of data or records)**

- **Destruction and deletion of critical assets**

- **Downloading from unauthorized sources or use of pirated software which might contain backdoors or malicious code**

- **Eavesdropping and packet sniffing**

- **Spoofing and impersonating other users**

- **Social engineering attacks**

- **Misuse of resources for non-business related or unauthorized activities**

- **Purposefully installing malicious software**


### Types of Insider Attackers

\begin{figure}
  \centering
  \href{https://www.varonis.com/blog/insider-threats/}{\includegraphics[height=0.78\textheight, width=\textwidth, keepaspectratio]{../../figures/insider-threats-types.png}}
\end{figure}

Source: Slide is based on @peters2019.

<!-- A legitimate user with proper access credentials gone rouge
◦ Full knowledge of systems & security policies -->

<!-- An attacker who has stolen/obtained and uses -->
<!-- credentials of a legitimate user -->


### Indicators of Insider Attackers

\begin{figure}
  \centering
  \href{https://www.varonis.com/blog/insider-threats/}{\includegraphics[height=0.78\textheight, width=\textwidth, keepaspectratio]{../../figures/insider-threats-indicators-1024x789.png}}
\end{figure}

Source: Slide is based on @peters2019.


### Insider Threats: Defense

- **Turncloaks:**  
  * Decoys/traps (e.g., honeypots)  

    
- **Pawns:**    
  * Behavioral profiling & anomaly detection  
  * Requires extensive logging of systems & users
  

**Note that the use of the aforementioned defense mechanisms may be suitable for the other type of attacker as well.**

### Types of Data Used for Insider Attack Detection

- **CLI command sequences** 

- **System calls** 

- **Database/file accesses** 

- **Keystroke dynamics** 

- **Mouse dynamics** 


\vspace{2\baselineskip}

Source: Slide is adapted from CS259D, Stanford University, \url{https://web.stanford.edu/class/cs259d/lectures/Session3.pdf}

### A Taxonomy of Modeling Approaches
\begin{figure}[h!]
		\centering
		\begin{tikzpicture}[node distance=2cm, font=\small]
		\tikzstyle{start} = [rectangle, rounded corners, minimum width=0.3\textwidth, minimum height=1cm,text badly centered, text width=0.3\textwidth, draw=black, thick,font=\normalfont]
		
		\tikzstyle{categories} = [rectangle, rounded corners, minimum width=0.3\textwidth, minimum height=1cm,text width=0.3\textwidth, draw=black, fill=gray!10, thick, font=\scriptsize]
		
		
		\tikzstyle{arrow} = [->,>= triangle 45]
		
		% Begining of the creation of the figure
		\node (start) [start] {\textbf{Modeling Approaches}};
		
		% Level 1 nodes
		\node (ia1) [categories, below of = start, xshift=-0.35\textwidth] {\textbf{\underline{Surveillance}} \\ (A) Statistical Approach \\ (B) Deviations from baseline \\ (C) Can capitalize on time};
		\node (ia2) [categories, below of = start, xshift = 0in] {\textbf{\underline{Single-class classification}} \\ (A) ML Approach \\ (B) Goal: Similar to Surveillance \\ (C) Cannot capitalize on time};	
		\node (ia3) [categories, below of = start, xshift=0.35\textwidth] {\textbf{\underline{Multi-class classification}} \\ (A) One user = one class \\ (B) Retraining w/ user changes \\ (C) Non-self samples??};
		
		% Arrows
		\draw [arrow] (start.south) -| (ia1);
		\draw [arrow] (start.south) -| (ia2);
		\draw [arrow] (start.south) -| (ia3);
		\end{tikzpicture}

\end{figure}

\centering{An overview of three possible modeling paradigms.}


### A Statistical Surveillance Approach: Bootstrap Sampling + Exceedance Probability Limits {.allowframebreaks}

\centering{\href{https://towardsdatascience.com/an-introduction-to-the-bootstrap-method-58bcb51b4d60}{\includegraphics[height=0.7\textheight, frame]{../../figures/bootstrap.png}}}

```{r bootstrap}
  B = 500 # number of bootstrap samples
  h = 'e2693' # user
  p = 90 #% # percentile used for threshold calculation
  ep = 95 # exceedance probablity
  df = read.csv("../../data/16-insiderThreat.csv", stringsAsFactors = F) 
  
  df %>% filter(UserAbbrv == h) -> dfUser
```

```{r bootstrap2}
  x = as.matrix(dfUser$Risk.Score)
  n = 100
  bootSamples = replicate(B, x[sample.int(n, replace=TRUE),] )
  
  BootstrapQuantiles = apply(bootSamples, MARGIN=2, FUN=quantile, prob= p/100)
  threshold = quantile(BootstrapQuantiles, probs = ep/100)
  threshold = round(threshold,0)
  
  print(threshold)

```

# Biometrics for UBA

### Human-Human Authentication

\begin{itemize}
\item \textbf{User authentication} 
\begin{itemize}
\item \textbf{Differentiate one human user from another}
\end{itemize}
\item \textbf{Prominent authentication approaches}
\begin{itemize}
\item \textbf{Passwords}
\item \textbf{Traditional biometrics}
\end{itemize}
\end{itemize}




### Limitations of Existing User Authentication Solutions [1]

\begin{columns}
\begin{column}{0.48\textwidth}
\begin{itemize}
\item \textbf{Passwords} 
\begin{itemize}
\item \textbf{Either insecure or unusable}
\end{itemize}
\item \textbf{Traditional biometrics (e.g., fingerprints)}
\begin{itemize}
\item \textbf{Invasive}
\item \textbf{High rejection rates}
\item \textbf{Require additional hardware}
\item \textbf{Susceptible to impersonation or spoofing} 
\end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\begin{center}
\includegraphics[width=\textwidth, height=0.8\textheight, frame, keepaspectratio]{../../figures/password_usability.jpg}
\end{center}
\end{column}
\end{columns}


### Limitations of Existing User Authentication Solutions [2]

\begin{columns}
\begin{column}{0.48\textwidth}
\begin{itemize}
\item \textbf{Passwords} 
\begin{itemize}
\item \textbf{Either insecure or unusable} 
\end{itemize}
\item \textbf{Traditional biometrics (e.g., fingerprints)}
\begin{itemize}
\item \textbf{Invasive}
\item \textbf{High rejection rates}
\item \textbf{Require additional hardware}
\item \textbf{Susceptible to impersonation or spoofing} 
\end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\begin{center}
\includegraphics[width=\textwidth, height=0.8\textheight, frame, keepaspectratio]{../../figures/spoofing.png}
\end{center}
\end{column}
\end{columns}


### Behavioral Biometrics

- **Keystroke dynamics @monrose1997authentication**  
- **Mouse movement patterns @zheng2011efficient**
- **Touch gesture biometrics**  
  - Sliding horizontally and vertically @frank2012touchalytics  
  - Sliding up, down, left,  and right and tap @li2013unobservable  
  - Horizontal slide and the pattern unlock @de2012touch

\begin{center}
\includegraphics[width=0.3\linewidth]{../../figures/bio2.jpg}
\end{center}


### Gametrics
\begin{itemize}
\item \textbf{Interactive game-based behavioral biometrics}
\item \textbf{Why games?}

  \begin{itemize}
  	\item \textbf{Fully supported by web browsers and touch screen devices}
  	\item \textbf{Randomized, interactive and cognitive nature}
  	\item \textbf{Sufficient cues can be extracted in a short period of time} 
  \end{itemize}

\end{itemize}



### Game Cognitive Task

\begin{center}
	\includemedia[width=0.6\linewidth,height=0.75\textheight,activate=pageopen,
	passcontext,
	transparent,
	addresource=game.mp4,
	flashvars={source=game.mp4}
	]{\includegraphics[width=0.6\linewidth, frame, height=0.75\textheight, keepaspectratio]{../../figures/game.PNG}}{VPlayer.swf}
	
\end{center}



### Features \& Classification Metrics
\begin{itemize}
\item \textbf{Features}
\begin{itemize}
\item \textbf{Mouse dynamics / touch gesture}  
\item \textbf{Cognitive ability}
\item \textbf{Others (i.e., distance-based features)}
\end{itemize}

\item \textbf{Classifier}
\begin{itemize}
\item \textbf{Random forest}
\end{itemize}

\item \textbf{Classification metrics} 
\begin{itemize}
\item \textbf{False Positive Rate: } Measures the security 
\item \textbf{False Negative Rate:} Measures the usability

\end{itemize}
\end{itemize}




### Inter- \& Intra- Session Analysis [1]
\begin{itemize}
\item \textbf{Web-based study (MTurk)}
\item \textbf{Data collection methodology:}
\begin{itemize}
\item Day 1: 98 participants -- 60 challenges
\item Day 2: 62 participants -- 36 challenges
\item Day 3: 29 participants -- 36 challenges
\end{itemize}

\item \textbf{Number of successfully completed challenges = 9076}
\item \textbf{Average solving time = 7.5s}
\end{itemize}



### Inter- \& Intra-Session Results [2]

\begin{center}
\includegraphics[width=0.9\linewidth, height=0.75\textheight, keepaspectratio]{../../figures/result1.png}
\end{center}



### Interactive Biometrics Discussion

\begin{itemize}
\item \textbf{Efficiency}
\begin{itemize}
\item \textbf{Short enrollment time}
\item \textbf{Short time to identify the user}
\item \textbf{Building and updating the classifier and testing a new instance take short time}
\end{itemize}

\item \textbf{Application Scenarios}
\begin{itemize}
\item \textbf{Point-of-entry}
\item \textbf{Integrated with graphical passwords}
\item \textbf{Fall-back authentication}
\end{itemize}

\end{itemize}


# Gait for Biometrics

### Experimental Data {.allowframebreaks}

\begin{columns}
    \begin{column}{0.3\textwidth}
        \begin{figure}[b]
            \centering
            \includegraphics[height=0.35\textheight]{../../figures/pa.jpg}
            \caption{Part assembly}
            \label{fig:pa}
        \end{figure}
    \end{column}
    \begin{column}{0.3\textwidth}
        \begin{figure}[b]
            \centering
            \includegraphics[height=0.35\textheight]{../../figures/si.jpg}
            \caption{Supply pickup}
            \label{fig:si}
        \end{figure}
    \end{column}
    \begin{column}{0.3\textwidth}
        \begin{figure}[b]
            \centering
            \includegraphics[height=0.35\textheight]{../../figures/mmh.jpg}
            \caption{MMH}
            \label{fig:mmh}
        \end{figure}
    \end{column}
\end{columns}

Figures are from @maman2017data.


\begin{center}
\includegraphics[width=\textwidth, height=0.7\textheight, keepaspectratio]{../../figures/tasksvideo.png}
\end{center}

Video is based on research presented in @maman2017data.


### Gait Cycle Analysis {.allowframebreaks}
\begin{figure}
	\centering
		\includegraphics[height=0.5\textheight, frame, trim={0 0.4cm 0 0}, clip]{../../figures/ankle.jpeg}
		\caption*{The relative size of the IMU when placed on the right ankle}
\end{figure} 
Figure is from research presented in @baghdadi2019monitoring.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth, height=0.8\textheight, keepaspectratio]{../../figures/IMUpreprocess.jpg}
\end{figure}

\begin{figure}
\fbox{
  \animategraphics[label= fig:step, controls, autoplay, width= 0.5\linewidth]{1}{../../figures/Step}{1}{10}
}
\end{figure}
Joint work with researchers at the University at Buffalo.

\begin{figure}
\fbox{
  \animategraphics[label= fig:cusum, controls, autoplay, width= 0.5\linewidth]{1}{../../figures/cusum}{1}{10}
}
\end{figure}
Based on research presented in @baghdadi2019monitoring.




# Recap


### Interactive Biometrics Limitations and Future Work

\begin{itemize}
\item \textbf{Study the effect of user's behavior variation on the accuracy}

\item \textbf{Test the accuracy when switching devices or hardware}

\item  \textbf{Add more complexity to the game challenges to increase the level of interaction, and improve the overall usability and security}

\end{itemize}


### What you should have learned today

\begin{block}{\textbf{Objectives}}
  \begin{itemize}
    \item \textbf{One class classifiers}
    \item \textbf{Describe different applications for UBA}
  \end{itemize}
\end{block}



### References {.allowframebreaks}
::: {#refs}
:::