---
title: "My First RMarkdown Report"
author: "Fadel M. Megahed"
date: "`r format(Sys.Date(), format = '%B %d, %Y')`"
output: 
  html_document:
    code_folding: show
    code_download: TRUE
    number_sections: TRUE
    paged_df: TRUE
    toc: TRUE
    toc_float: TRUE
    theme: readable
params:
  # the parameters for the report is optional and will be often ignored in most
  # RMarkdown examples you see online; however, it is handy if you will repeat
  # your analysis for different settings
  lines_to_skip: 1099999
  rows_to_read: 100000
---



```{r setup, include=FALSE}
# setup chunk will not be included since include = FALSE
knitr::opts_chunk$set(echo = TRUE) # default show code and output
```


---


# Required Libraries

Insert all the R packages that you have used here.
```{r packages}
# libraries
if(require(reticulate)==FALSE) install.packages('reticulate')
library(reticulate)

if(require(tidyverse)==FALSE) install.packages('tidyverse')
library(tidyverse)
```


---

# Assignment Questions

In this assignment, you will *knit* this Markdown and generate a report that will explore the [LANL Dataset](https://www.dropbox.com/s/88ajdyp7hfnvrl2/lanl-auth-dataset-1-03.csv), whose documentation is available at [User-Computer Authentication Associations in Time - Cyber Security Research](https://csr.lanl.gov/data/auth/).

**For the data in lines 1.1M-1.2M, please answer the following questions:**  

- What is the timestamp of observation (i.e. row number) 1,200,000?    

- What are the number of unique users for the aforementioned 100,000 observations?   

- What is the number of unique computers used by U12 for the aforementioned 100,000 observations?  

---

# R Output

Insert some text here describing how you read the data set and the steps taken to answer the three questions

```{r lanl_tidyverse}
# extracting the parameters and making them  R objects
lines_to_skip = params$lines_to_skip
rows_to_read = params$rows_to_read


```

---

# Python Output

Talk about the Python code here

```{python lanl_pandas}
import pandas as pd

# converting the R object into a python object
lines_to_skip = r.lines_to_skip
rows_to_read = r.rows_to_read


```
